Symmetry, stability, and computation of degenerate lasing modes

### 2017-01-10 degeneracy_paper.tex
fixed references (Number --> Pages for Phys Rev)
fixed single error with year of Abraham paper
added notes on how to convert from Lyx to Tex for submission to pRA
finished referee changes

@@

\begin{align}
-\varepsilon\ddot{\mathbf{E}}^{+} & =\nabla\times\nabla\times\mathbf{E}^{+}+\ddot{\mathbf{P}}^{+}\nonumber \\
i\dot{\mathbf{P}}^{+} & =(\omega_{a}-i\gamma_{\perp})\mathbf{P}^{+}+\gamma_{\perp}\mathbf{E}^{+}D\label{eq:maxwell--bloch}\\
\dot{D} & =\gamma_{\parallel}(D_{0}-D)+\mathrm{Im}(\mathbf{E}^{-}\cdot\mathbf{P}^{+}),\nonumber 
\end{align}
where $\mathbf{E}^{+}(\mathbf{x},t)$ is the ``positive-frequency''
component of the electric field {[}with $\mathbf{E}^{-}=\mathbf{E}^{+\star}$
and the physical field given by $2\mathrm{Re}(\mathbf{E}^{+})${]},
$\varepsilon(\mathbf{x})$ is the ``cold-cavity'' permittivity (not
including the gain transition), $\mathbf{P}(\mathbf{x},t)$ is the
polarization describing a transition (of frequency $\omega_{a}$ and
linewidth $\gamma_{\perp}$) between two atomic energy levels, $D(\mathbf{x},t)$
is the population inversion between those two levels (with relaxation
rate $\Gamma_{\Vert}$), and $D_{0}(\mathbf{x})$ is the strength
of a pumping process driving the inversion. Additionally, for convenience,
one chooses units such that the following factors are set to unity:
the dipole moment matrix element of the two level system, Planck's
constant $\hbar$, and the speed of light $c$. Using the stationary
inversion approximation $D(\mathbf{x},t)\approx D(\mathbf{x})$ \cite{fu_multifrequency_1991,tureci_self-consistent_2006}
along with an ansatz of a finite number of lasing modes
\begin{equation}
\mathbf{E}^{+}(\mathbf{x},t)=\sum_{\nu}\mathbf{E}_{\nu}(\mathbf{x})e^{-i\omega_{\nu}t},\label{eq:positive frequency component}
\end{equation}
where $\omega_{\nu}$ are the real mode frequencies, the second equation
in Eq. (\ref{eq:maxwell--bloch}) is solved to eliminate $\mathbf{P}^{+}$
as an unknown, and the third equation becomes
\begin{equation}
\dot{D}=\gamma_{\parallel}(D_{0}-D)+D\,\Im\left[\sum_{\mu\nu}\Gamma(\omega_{\nu})\mathbf{E}_{\mu}^{\star}\cdot\mathbf{E}_{\nu}e^{i(\omega_{\mu}-\omega_{\nu})t}\right],\label{eq:Ddot MB equation}
\end{equation}
where $\Gamma(\omega_{\mu})\equiv\gamma_{\perp}/(\omega_{\mu}-\omega_{a}+i\gamma_{\perp})$.
In order for the stationary inversion approximation $\dot{D}=0$ to
be valid, the oscillating terms on the right-hand side of Eq.~(\ref{eq:Ddot MB equation})
must average to zero on a timescale much faster than the relaxation
timescale $1/\gamma_{\parallel}$; and in order to do so, the beating
frequencies $\omega_{\mu}-\omega_{\nu}$ must be either exactly zero
or much faster than the relaxation rate $\gamma_{\parallel}$ \cite{fu_multifrequency_1991,tureci_self-consistent_2006,ge_steady-state_2010}.
This condition is evidently true when any pair of modes is either
exactly degenerate, or situated very far apart from each other in
frequency space. When these conditions are met, Eq. (\ref{eq:maxwell--bloch})
reduces to the SALT equation \cite{ge_steady-state_2010,tureci_self-consistent_2006,tureci_ab_2009}

@@

So far, the discussion and examples in this paper have dealt with
the case in which only one pair of degenerate modes are lasing. The
generalization to the case of multimode lasing (i.e. multiple nondegenerate
and degenerate lasing modes all lasing simultaneously) is straightforward.
Since our method combines degenerate pairs into a single mode that
is the stable linear combination (as given by the perturbation theory
in Sec.~\ref{sec:Threshold-perturbation-theory}, the multimode treatment
is exactly the same as for SALT without degeneracies: the degenerate
pairs are always treated as a single mode. As in previous work on
SALT \cite{ge_steady-state_2010,tureci_self-consistent_2006,direct_salt},
all lasing modes are solved simultaneously at first, and there the
collective effect of their spatial hole-burning is used to track the
passive modes and add any mode (degenerate or nondegenerate) that
crosses threshold to the list of lasing modes. The only aspects of
our method requiring generalization are the threshold perturbation
theory of Sec.~\ref{sec:Threshold-perturbation-theory} and the \emph{quadratic
program} \cite{boyd} (QP) method of Sec.~\ref{sub:Forcing-the-degeneracy}.
For both aspects, we describe small tweaks to the methods presented
in those sections that make them valid for the case of multimode lasing.

@@

  {volume} {92}} (\bibinfo {year} {2015})}\BibitemShut {NoStop}%

@@

  {Cerjan}, \citenamefont {Makris}, \citenamefont {Stone}, \citenamefont
  {Melenk}, \citenamefont {Johnson},\ and\ \citenamefont

@@

  {\bibfnamefont {K.~G.}\ \bibnamefont {Makris}}, \bibinfo {author}
  {\bibfnamefont {J.~M.}\ \bibnamefont {Melenk}}, \bibinfo {author}
  {\bibfnamefont {S.~G.}\ \bibnamefont {Johnson}}, \ and\ \bibinfo {author}
  {\bibfnamefont {S.}~\bibnamefont {Rotter}},\ }\href@noop {} {\bibfield
  {90}},\ \bibinfo {pages} {023816} (\bibinfo {year} {2014})}\BibitemShut

@@

\bibitem [{\citenamefont {Zeghlache}\ \emph {et~al.}(1987)\citenamefont

@@

  {37}} (\bibinfo {year} {1987})}\BibitemShut {NoStop}%

@@

  {Phys. Rev. A}\ }\textbf {\bibinfo {volume} {34}} (\bibinfo {year}
  {1986})}\BibitemShut {NoStop}%

@@

  {volume} {64}} (\bibinfo {year} {2001})}\BibitemShut {NoStop}%

@@

  {volume} {65}} (\bibinfo {year} {2002})}\BibitemShut {NoStop}%

@@

  {91}} (\bibinfo {year} {2015})}\BibitemShut {NoStop}%

@@

  }\textbf {\bibinfo {volume} {83}} (\bibinfo {year} {2011})}\BibitemShut

@@

  {Phys. Rev. E}\ }\textbf {\bibinfo {volume} {71}} (\bibinfo {year}
  {2005})}\BibitemShut {NoStop}%

>>>>

\selectlanguage{english}%

@@

\begin{align}
-\varepsilon\ddot{\mathbf{E}}^{+} & =\nabla\times\nabla\times\mathbf{E}^{+}+\ddot{\mathbf{P}}^{+}\nonumber \\
i\dot{\mathbf{P}}^{+} & =(\omega_{a}-i\gamma_{\perp})\mathbf{P}^{+}+\gamma_{\perp}\mathbf{E}^{+}D\label{eq:maxwell--bloch}\\
\dot{D} & =\gamma_{\parallel}(D_{0}-D)+\mathrm{Im}(\mathbf{E}^{-}\cdot\mathbf{P}^{+}),\nonumber 
\end{align}
where $\mathbf{E}^{+}(\mathbf{x},t)$ is the ``positive-frequency''
component of the electric field {[}with $\mathbf{E}^{-}=\mathbf{E}^{+\star}$
and the physical field given by $2\mathrm{Re}(\mathbf{E}^{+})${]},
$\varepsilon(\mathbf{x})$ is the ``cold-cavity'' permittivity (not
including the gain transition), $\mathbf{P}(\mathbf{x},t)$ is the
polarization describing a transition (of frequency $\omega_{a}$ and
linewidth $\gamma_{\perp}$) between two atomic energy levels, $D(\mathbf{x},t)$
is the population inversion between those two levels (with relaxation
rate $\Gamma_{\Vert}$), and $D_{0}(\mathbf{x})$ is the strength
of a pumping process driving the inversion. Additionally, for convenience,
one chooses units such that the following factors are set to unity:
the dipole moment matrix element of the two level system, Planck's
constant $\hbar$, and the speed of light $c$. Using the stationary
inversion approximation $D(\mathbf{x},t)\approx D(\mathbf{x})$ \cite{fu_multifrequency_1991,tureci_self-consistent_2006}
along with an ansatz of a finite number of lasing modes
\begin{equation}
\mathbf{E}^{+}(\mathbf{x},t)=\sum_{\nu}\mathbf{E}_{\nu}(\mathbf{x})e^{-i\omega_{\nu}t},\label{eq:positive frequency component}
\end{equation}
where $\omega_{\nu}$ are the real mode frequencies, the second equation
in Eq. (\ref{eq:maxwell--bloch}) is solved to eliminate $\mathbf{P}^{+}$
as an unknown, and the third equation becomes
\begin{equation}
\dot{D}=\gamma_{\parallel}(D_{0}-D)+D\,\Im\left[\sum_{\mu\nu}\Gamma(\omega_{\nu})\mathbf{E}_{\mu}^{\star}\cdot\mathbf{E}_{\nu}e^{i(\omega_{\mu}-\omega_{\nu})t}\right],\label{eq:Ddot MB equation}
\end{equation}
where $\Gamma(\omega_{\mu})\equiv\gamma_{\perp}/(\omega_{\mu}-\omega_{a}+i\gamma_{\perp})$.
In order for the stationary inversion approximation $\dot{D}=0$ to
be valid, the oscillating terms on the right-hand side of Eq.~(\ref{eq:Ddot MB equation})
must average to zero on a timescale much faster than the relaxation
timescale $1/\gamma_{\parallel}$. In order to do so, the beating
frequencies $\omega_{\mu}-\omega_{\nu}$ must be either exactly zero
or much faster than the relaxation rate $\gamma_{\parallel}$ \cite{fu_multifrequency_1991,tureci_self-consistent_2006,ge_quantitative_2008};
that is, two modes is either exactly degenerate or situated very far
apart from each other in frequency space, with the latter case resulting
in the time-dependent beating component of the inversion having a
negligible amplitude compared to the stationary component \cite{haken_light:_1985}.
When these conditions are met, Eq. (\ref{eq:maxwell--bloch}) reduces
to the SALT equation \cite{ge_steady-state_2010,tureci_self-consistent_2006,tureci_ab_2009}

@@

So far, the discussion and examples in this paper have dealt with
the case in which only one pair of degenerate modes are lasing. The
generalization to the case of multimode lasing (i.e. multiple nondegenerate
and degenerate lasing modes all lasing simultaneously) is straightforward.
Since our method combines degenerate pairs into a single mode that
is the stable linear combination (as given by the perturbation theory
in Sec.~\ref{sec:Threshold-perturbation-theory}, the multimode treatment
is exactly the same as for SALT without degeneracies: the degenerate
pairs are always treated as a single mode. As in previous work on
SALT \cite{ge_steady-state_2010,tureci_self-consistent_2006,direct_salt},
all lasing modes are solved simultaneously at first, and there the
collective effect of their spatial hole-burning is used to track the
passive modes and add any mode (degenerate or nondegenerate) that
crosses threshold to the list of lasing modes. As in the case of non-degenerate
SALT, the spacing between modes with different frequencies must remain
much larger than $\gamma_{\parallel}$ in order for the stationary
inversion approximation to remain valid (usually, lasing frequencies
do not appreciably deviate from their threshold values, so this condition
is often safely satisfied). The only aspects of our method requiring
generalization are the threshold perturbation theory of Sec.~\ref{sec:Threshold-perturbation-theory}
and the \emph{quadratic program} \cite{boyd} (QP) method of Sec.~\ref{sub:Forcing-the-degeneracy}.
For both aspects, we describe small tweaks to the methods presented
in those sections that make them valid for the case of multimode lasing.

@@

  {volume} {92}},\ \bibinfo {pages} {013847} (\bibinfo {year}
  {2015})}\BibitemShut {NoStop}%

@@

  {Cerjan}, \citenamefont {Ge}, \citenamefont {Makris}, \citenamefont {Stone},
  \citenamefont {Melenk}, \citenamefont {Johnson},\ and\ \citenamefont

@@

  {\bibfnamefont {L.}~\bibnamefont {Ge}}, \bibinfo {author} {\bibfnamefont
  {K.~G.}\ \bibnamefont {Makris}}, \bibinfo {author} {\bibfnamefont {A.~D.}\
  \bibnamefont {Stone}}, \bibinfo {author} {\bibfnamefont {J.~M.}\ \bibnamefont
  {Melenk}}, \bibinfo {author} {\bibfnamefont {S.~G.}\ \bibnamefont {Johnson}},
  \ and\ \bibinfo {author} {\bibfnamefont {S.}~\bibnamefont {Rotter}},\
  }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {Phys. Rev. A}\
  }\textbf {\bibinfo {volume} {90}},\ \bibinfo {pages} {023816} (\bibinfo
  {year} {2014})}\BibitemShut {NoStop}%

@@

\bibitem [{\citenamefont {Zeghlache}\ \emph {et~al.}(1988)\citenamefont

@@

  {37}},\ \bibinfo {pages} {470} (\bibinfo {year} {1988})}\BibitemShut

@@

  {Phys. Rev. A}\ }\textbf {\bibinfo {volume} {34}},\ \bibinfo {pages} {3101}
  (\bibinfo {year} {1986})}\BibitemShut {NoStop}%

@@

  {volume} {64}},\ \bibinfo {pages} {125122} (\bibinfo {year}
  {2001})}\BibitemShut {NoStop}%

@@

  {volume} {65}},\ \bibinfo {pages} {195306} (\bibinfo {year}

@@

  {91}},\ \bibinfo {pages} {063806} (\bibinfo {year} {2015})}\BibitemShut

@@

  }\textbf {\bibinfo {volume} {83}},\ \bibinfo {pages} {205131} (\bibinfo
  {year} {2011})}\BibitemShut {NoStop}%

@@

  {Phys. Rev. E}\ }\textbf {\bibinfo {volume} {71}},\ \bibinfo {pages} {026602}
  (\bibinfo {year} {2005})}\BibitemShut {NoStop}%
